---
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Probability and Statistics

# Lab Assignment 1: Naive Bayes Classifier

### *Name1 Surname1, Name2 Surname2, Name3 Surname3*

## Introduction

During the past three weeks, you learned a couple of essential notions
and theorems, and one of the most important among them is the *Bayes
theorem*.

One of its applications is **Naive Bayes classifier**, which is a
probabilistic classifier whose aim is to determine which class some
observation probably belongs to by using the Bayes formula:
$$\mathsf{P}(\mathrm{class}\mid \mathrm{observation})=\frac{\mathsf{P}(\mathrm{observation}\mid\mathrm{class})\mathsf{P}(\mathrm{class})}{\mathsf{P}(\mathrm{observation})}$$

Under the strong independence assumption, one can calculate
$\mathsf{P}(\mathrm{observation} \mid \mathrm{class})$ as
$$\mathsf{P}(\mathrm{observation}) = \prod_{i=1}^{n} \mathsf{P}(\mathrm{feature}_i), \qquad \mathsf{P}(\mathrm{observation} \mid \mathrm{class}) = \prod_{i=1}^{n} \mathsf{P}(\mathrm{feature}_i \mid \mathrm{class}),$$
where $n$ is the total number of features describing a given
observation. Thus, $\mathsf{P}(\mathrm{class}|\mathrm{observation})$ now
can be calculated as

$$\mathsf{P}(\mathrm{class} \mid \mathrm{\mathrm{observation}}) = \mathsf{P}(\mathrm{class})\times \prod_{i=1}^{n}\frac{\mathsf{P}(\mathrm{feature}_i\mid \mathrm{class})}{\mathsf{P}(\mathrm{feature}_i)}\tag{1}$$

All the terms on the right-hand side can be estimated from the data as
respective relative frequencies;\
see [this
site](https://monkeylearn.com/blog/practical-explanation-naive-bayes-classifier/)
for more detailed explanations.

## Data description

There are 5 datasets uploaded on the cms.

To determine your variant, take your team number from the list of teams
on cms and take *mod 5* - this is the number of your data set.

-   **0 - authors** This data set consists of citations of three famous
    writers: Edgar Alan Poe, Mary Wollstonecraft Shelley and HP
    Lovecraft. The task with this data set is to classify a piece of
    text with the author who was more likely to write it.

-   **1 - discrimination** This data set consists of tweets that have
    discriminatory (sexism or racism) messages or of tweets that are of
    neutral mood. The task is to determine whether a given tweet has
    discriminatory mood or does not.

-   **2 - fake news** This data set contains data of American news: a
    headline and an abstract of the article. Each piece of news is
    classified as fake or credible. The task is to classify the news
    from test.csv as credible or fake.

-   **3 - sentiment** All the text messages contained in this data set
    are labeled with three sentiments: positive, neutral or negative.
    The task is to classify some text message as the one of positive
    mood, negative or neutral.

-   **4 - spam** This last data set contains SMS messages classified as
    spam or non-spam (ham in the data set). The task is to determine
    whether a given message is spam or non-spam.

Each data set consists of two files: *train.csv* and *test.csv*. The
first one you will need find the probabilities distributions for each of
the features, while the second one is needed for checking how well your
classifier works.

```{r}
# here goes a list of recommended libraries,
# though you may install other ones if they are needed
library(tidytext)
library(readr)
library(dplyr)
library(ggplot2)
```

## Instructions

-   The first step is data pre-processing, which includes removing
    punctuation marks and stop words

-   represent each message as a bag-of-words

-   using the training set, calculate all the conditional probabilities
    in formula (1)

-   use those to predict classes for messages in the test set

-   evaluate effectiveness of the classifier by calculating the
    corresponding metrics

-   shortly summarize your work

-   do not forget to submit both the (compiled) Rmd source file and the .html
    output

### Data pre-processing

-   Read the *.csv* data files.
-   Сlear your data from punctuation or other unneeded symbols.
-   Clear you data from stop words. You don't want words as is, and, or
    etc. to affect your probabilities distributions, so it is a wise
    decision to get rid of them. Find list of stop words in the cms
    under the lab task.
-   Represent each test message as its bag-of-words. Here:
    <https://machinelearningmastery.com/gentle-introduction-bag-words-model/>
    you can find general introduction to the bag-of-words model and
    examples on to create it.
-   It is highly recommended to get familiar with R dataframes, it would
    make the work much easier to do.
-   Useful links:
    -   <https://steviep42.github.io/webscraping/book/bagofwords.html#tidytext> -
        example of using *tidytext* to count frequencies of the words.
    -   Basics of Text Mining in R:
        <http://rstudio-pubs-static.s3.amazonaws.com/256588_57b585da6c054349825cba46685d8464.html>
        . Note that it also includes an example on how to create a bag
        of words from your text document.

```{r}
list.files(getwd())
list.files("data/4-spam")
```

```{r}
test_path <- "data/4-spam/test.csv"
train_path <- "data/4-spam/train.csv"

stop_words <- read_file("stop_words.txt")
splitted_stop_words <- strsplit(stop_words, split='\n')
splitted_stop_words <- splitted_stop_words[[1]]
```

```{r}
train <-  read.csv(file = train_path, stringsAsFactors = FALSE)
test <-  read.csv(file = test_path, stringsAsFactors = FALSE)
```

```{r}
# note the power functional features of R bring us! 
tidy_text <- unnest_tokens(train, output = 'word', input ='Message') %>%
             filter(!word %in% splitted_stop_words)

df <- tidy_text %>% count(word, sort=TRUE)

df_with_categories <- tidy_text %>% count(Category, word, sort=TRUE)
spam_df <- df_with_categories %>% filter(Category=="spam")
non_spam_df <- df_with_categories %>% filter(Category=="ham")
```

### Data visualization

Each time you work with some data, you need to understand it before you
start processing it. R has very powerful tools to make nice plots and
visualization. Show what are the most common words for negative and
positive examples as a histogram, word cloud etc. Be creative!


## Classifier implementation

```{r}
spam_test_df = test %>% filter(Category=="spam") %>% count(Category, Message, result=1)
ham_test_df = test %>% filter(Category=="ham") %>% count(Category, Message, result=1)
is_spam = nrow(train %>% filter(Category=="spam"))/nrow(train)
is_not_spam = 1 - is_spam
```

```{r}
naiveBayes <- setRefClass("naiveBayes",

       # here it would be wise to have some vars to store intermediate result
       # frequency dict etc. Though pay attention to bag of words!

        # X_test - for spam, y_test - for ham (non-spam)
       fields = list(X_test="matrix", y_test="matrix"),

       methods = list(
                    # prepare your training data as X - bag of words for each of your
                    # messages and corresponding label for the message encoded as 0 or 1
                    # (binary classification task)

                    fit = function(X, y)
                    {
                      for (row in 1:nrow(spam_test_df)){
                        val = predict(spam_test_df[row, -4])
                        if (val == 1){
                          spam_test_df[row, "result"] = 0
                        }
                      }
                      for (row in 1:nrow(ham_test_df)){
                        val = predict(ham_test_df[row, -4])
                        if (val == 0){
                            ham_test_df[row, "result"] = 0
                        }
                      }
                        X_test <<- spam_test_df
                        y_test <<- ham_test_df
                    },

                    # return prediction for a single message

                    # 0 - spam, 1 - non-spam
                    predict = function(message)
                    {
                         #TODO: split message into words, for each list calc the probability, return the highest
                      words <- unnest_tokens(message, output = 'word', input = 'Message')
                      dfs = c(spam_df, non_spam_df)
                      res = c(0, 0)
                      for (ind in 1:2){
                        df = dfs[ind]
                        sum = 1
                        for (row in 1:nrow(words)) {
                            sum = sum * calc_probability(words[row, "word"], df)
                        }
                        if (ind == 1){
                          res[1] = sum * is_spam
                        }
                        else{
                          res[2] = sum * is_not_spam
                        }
                      }
                      if (res[1] > res[2]){
                        return(0)
                      }
                      return(1)
                    },

                    # score you test set so to get the understanding how well you model
                    # works.
                    # look at f1 score or precision and recall
                    # visualize them
                    # try how well your model generalizes to real world data!
                    score = function(X_test, y_test)
                    {
                         # TODO
                    }
))

model = naiveBayes()
model$predict(spam_test_df[1, -4])
model$fit()
```

## Measure effectiveness of your classifier
-   Note that accuracy is not always a good metric for your classifier.
    Look at precision and recall curves, F1 score metric.
-   Visualize them.
-   Show failure cases.

## Conclusions

Summarize your work by explaining in a few sentences the points listed
below.

-   Describe the method implemented in general. Show what are
    mathematical foundations you are basing your solution on.
-   List pros and cons of the method. This should include the
    limitations of your method, all the assumption you make about the
    nature of your data etc.
